title: "Running SSE with restart, plotting, and uncertainty-driven parameter sources"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running SSE with restart, plotting, and uncertainty-driven parameter sources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
can_run <- identical(Sys.getenv("RUN_NLMIXR2SSE_VIGNETTE"), "true")
can_plot <- can_run &&
  requireNamespace("tidyr", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
```

`nlmixr2sse` implements a PsN-style stochastic simulation and estimation
workflow for `nlmixr2`. Phase 6 covers three simulation modes plus resumable
run directories and summary recomputation, and Phase 7 adds plotting helpers
plus a simulated-data post-processing hook:

- fixed point estimates from the reference fit
- theta draws from `fit$cov`
- canonical raw-results input, typically from a bootstrap or SIR run
- rerunning the same output directory to resume or extend a completed study

## Setup

```{r setup, message = FALSE, warning = FALSE}
library(nlmixr2)
library(nlmixr2data)
library(nlmixr2sse)

set.seed(42)
samples <- 5L
```

## Reference and alternative models

```{r models}
ref_model <- function() {
  ini({
    tka <- log(1.57)
    tcl <- log(2.72)
    tv <- log(31.5)
    eta.ka ~ 0.6
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl)
    v <- exp(tv)
    cp <- linCmt()
    cp ~ add(add.sd)
  })
}

alt_model <- function() {
  ini({
    tka <- log(1.57)
    tcl <- log(2.72)
    tv <- log(31.5)
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl)
    v <- exp(tv)
    cp <- linCmt()
    cp ~ add(add.sd)
  })
}
```

```{r fit-models, message = FALSE, warning = FALSE, eval = can_run}
fit <- nlmixr2utils::nlmixr2(
  ref_model,
  nlmixr2data::theo_sd,
  est = "focei",
  control = list(print = 0L, covMethod = "r")
)

alt_fit <- nlmixr2utils::nlmixr2(
  alt_model,
  nlmixr2data::theo_sd,
  est = "focei",
  control = list(print = 0L, covMethod = "r")
)
```

## Fixed-parameter SSE

```{r run-sse, message = FALSE, warning = FALSE, eval = can_run}
sse <- runSSE(
  fit,
  alternativeModels = sseModel(alt_fit, label = "no_eta"),
  samples = samples,
  seed = 42,
  control = runSSEControl(workers = 1L)
)
```

```{r summary, eval = can_run}
summary(sse)
head(sse$parameterSummary)
head(sse$ofvSummary)
```

```{r tidy-wide, eval = can_plot}
tidyr::pivot_wider(
  sse$parameterSummary,
  id_cols = c(model_label, role, parameter, matched),
  names_from = statistic,
  values_from = value
)
```

```{r phase7-plots, eval = can_plot}
plotSSEParameterBias(sse) +
  ggplot2::theme(legend.position = "bottom")

plotSSEOfvDistribution(sse)
```

## Power and Type I summaries

`ofvSummary` reports `delta_OFV = reference - alternative`, so large positive
values support the alternative and large negative values support the reference.
The default thresholds are `0`, `3.84`, `5.99`, `7.81`, `9.49`, and `10.83`.
`pct_delta_above` is the Type I direction; `pct_delta_below` is the power
direction.

## Restart, add models, and recompute

`runSSE()` writes a numbered or explicit run directory with `run_info.rds`,
`sse_state.rds`, `initial_estimates.csv`, canonical `raw_results.*`, and the
summary artifacts. Calling `runSSE()` again on the same directory with
`restart = FALSE` resumes incomplete work or reloads a completed run instead of
refitting everything.

```{r phase6-run-dir, eval = can_run}
phase6_dir <- tempfile("nlmixr2sse-phase6-")

sse_phase6 <- runSSE(
  fit,
  alternativeModels = sseModel(alt_fit, label = "no_eta"),
  samples = samples,
  seed = 300,
  control = runSSEControl(workers = 1L),
  outputDir = phase6_dir,
  restart = TRUE
)

sse_resumed <- runSSE(
  fit,
  alternativeModels = sseModel(alt_fit, label = "no_eta"),
  samples = samples,
  seed = 999,
  control = runSSEControl(workers = 1L),
  outputDir = phase6_dir,
  restart = FALSE
)

nrow(sse_resumed$rawResults)
```

If you later want to evaluate another alternative on the exact same simulated
datasets, `addModels = TRUE` fits only the newly supplied labels. The example
below reuses the same fitted alternative model with a new label just to
demonstrate the workflow mechanics.

```{r phase6-add-models, message = FALSE, warning = FALSE, eval = can_run}
sse_added <- runSSE(
  fit,
  alternativeModels = sseModel(alt_fit, label = "no_eta_copy"),
  samples = samples,
  control = runSSEControl(addModels = TRUE, workers = 1L),
  outputDir = phase6_dir,
  restart = FALSE
)

sort(unique(sse_added$rawResults$model_label))
```

`recomputeSSE()` rebuilds `sse_results` and `sse_summary` outputs from the
saved raw results and initial estimates without rerunning simulations or fits.
Use this when you want a different `outFilter`, `refOfv`, or a subset of the
completed replicates.

```{r phase6-recompute, eval = can_run}
sse_recomputed <- recomputeSSE(
  phase6_dir,
  outFilter = ~ minimization_successful == 1,
  overwrite = FALSE
)

summary(sse_recomputed)
```

## Uncertainty modes

`parameterSource = "covariance"` draws theta values from `fit$cov` for each
replicate, and also draws OMEGA -- from a mean-centred inverse-Wishart per
block by default, or jointly with theta on a log-Cholesky scale when
`covarianceDraw = "joint"`. Residual error is drawn with the thetas because
`nlmixr2` parameterizes it as ordinary theta values, not as a separate SIGMA
block. An OMEGA block is drawn only when `fit$cov` covers all of its declared
entries; otherwise that block stays fixed at its fitted value while covered
thetas are still drawn.

```{r covariance-example, message = FALSE, warning = FALSE, eval = can_run}
sse_cov <- runSSE(
  fit,
  alternativeModels = sseModel(alt_fit, label = "no_eta"),
  samples = samples,
  seed = 100,
  control = runSSEControl(
    parameterSource = "covariance",
    workers = 1L
  )
)
```

For uncertainty taken from an external bootstrap rather than the asymptotic
covariance, use `parameterSource = "rawres"` with a canonical raw-results
file. The package ships a tiny bootstrap-style fixture so you can see the
cross-package interface without running a bootstrap first.

```{r rawres-path}
bootstrap_rawres <- system.file(
  "extdata",
  "bootstrap_raw_results.csv",
  package = "nlmixr2sse"
)
bootstrap_rawres
```

```{r rawres-example, message = FALSE, warning = FALSE, eval = can_run}
sse_rawres <- runSSE(
  fit,
  alternativeModels = sseModel(alt_fit, label = "no_eta"),
  samples = 2L,
  seed = 200,
  control = runSSEControl(
    parameterSource = "rawres",
    rawresInput = bootstrap_rawres,
    inFilter = ~ minimization_successful == 1,
    workers = 1L
  )
)
```

## Simulation post-processing

Use `simulationPostProcess` when the raw simulated output needs a final R-side
adjustment before refitting. The hook receives the merged estimation data plus
context like `sample`, `paramSet`, `solved`, and `referenceData`, and it must
return a data frame.

```{r phase7-postprocess, eval = can_run}
sse_post <- runSSE(
  fit,
  alternativeModels = sseModel(alt_fit, label = "no_eta"),
  samples = 2L,
  seed = 250,
  control = runSSEControl(
    workers = 1L,
    simulationPostProcess = function(data, sample) {
      data$SIM_REPLICATE <- sample
      data
    }
  )
)
```

Set `RUN_NLMIXR2SSE_VIGNETTE=true` before rendering if you want the fit and
`runSSE()` chunks to execute during vignette build.
