---
title: "How many subjects do I need to detect a covariate effect?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How many subjects do I need to detect a covariate effect?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
example_path <- system.file(
  "extdata",
  "sse_power_example.rds",
  package = "nlmixr2sse"
)
if (!nzchar(example_path)) {
  example_path <- normalizePath(
    file.path("..", "inst", "extdata", "sse_power_example.rds"),
    mustWork = TRUE
  )
}
example_data <- readRDS(example_path)
can_regenerate <- identical(Sys.getenv("RUN_NLMIXR2SSE_VIGNETTE"), "true")
```

Power calculations for covariate effects in nonlinear mixed-effects models are
usually not available in closed form. A practical alternative is stochastic
simulation and estimation (SSE): simulate data under a known truth, fit the
models you would compare in practice, and record how often the effect is
detected.

This vignette shows that workflow for a deliberately small example. The default
build path uses a precomputed result object shipped with the package, so the
article stays lightweight while still showing the full analysis pattern.

## Scientific Scenario

Suppose a single-dose oral PK study compares two weight groups in healthy
volunteers. We assume the true population has a weight effect on clearance and
ask a simple design question:

> How many subjects do we need to detect that clearance is higher in the
> high-weight group?

To keep the mechanics front and center, the example uses:

- a one-compartment model with first-order absorption;
- a binary covariate on clearance;
- a small grid of study sizes;
- likelihood-ratio testing between nested models.

## Data-Generating and Analysis Models

The data-generating model includes a positive covariate effect on `log(CL)`.
For each simulated data set, we fit:

- a base model without the weight effect; and
- a full model with the weight effect on clearance.

In a live study, that setup could look like this:

```{r model-code, eval = FALSE}
library(nlmixr2)
library(nlmixr2sse)

true_model <- function() {
  ini({
    tka <- log(1.2)
    tcl <- log(4.5)
    tv <- log(45)
    tcl_wt <- 0.30
    eta.cl ~ 0.09
    eta.v ~ 0.09
    prop.sd <- 0.20
  })
  model({
    wt_high <- WTGRP
    ka <- exp(tka)
    cl <- exp(tcl + tcl_wt * wt_high + eta.cl)
    v <- exp(tv + eta.v)
    cp <- linCmt()
    cp ~ prop(prop.sd)
  })
}
```

The shipped vignette does not rerun those fits by default. Instead, it loads a
small precomputed artifact that stores the resulting power summary and a set of
replicate-level covariate estimates for one study size.

```{r artifact-summary}
str(example_data, max.level = 1)
```

## The SSE Loop

For each study size in the grid, the conceptual loop is:

1. Simulate many replicate data sets under the true model.
2. Fit the base model and full model to each replicate.
3. Compute `delta OFV = OFV_base - OFV_full`.
4. Mark the covariate as detected when the likelihood-ratio test crosses the
   chosen threshold.
5. Tally convergence and summarize the covariate estimates.

The default vignette artifact uses the following study-size grid:

```{r study-grid}
example_data$study_sizes
```

Maintainers can regenerate the shipped object by setting
`RUN_NLMIXR2SSE_VIGNETTE=true` and replacing the saved artifact with a refreshed
small example run.

```{r regenerate-note, eval = FALSE}
if (can_regenerate) {
  # Run a small maintainer-only SSE workflow here and save a refreshed
  # inst/extdata/sse_power_example.rds artifact.
}
```

## Empirical Power Across Study Sizes

The precomputed power summary shows the typical pattern: power rises as the
study includes more subjects.

```{r power-table}
power_summary <- example_data$power_curve
power_summary
```

```{r fig-power, fig.width = 6, fig.height = 4}
library(ggplot2)

ggplot(
  power_summary,
  aes(x = study_size, y = 100 * power)
) +
  geom_line(linewidth = 0.9, colour = "#2f5c85") +
  geom_point(size = 2.2, colour = "#2f5c85") +
  geom_hline(yintercept = 80, linetype = 2, colour = "#b24a2a") +
  scale_x_continuous(breaks = power_summary$study_size) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    x = "Subjects",
    y = "Empirical power (%)",
    title = "Power increases with study size"
  ) +
  theme_minimal()
```

The dashed line marks the common 80% benchmark. In this example, the power
curve suggests that the design starts to approach that threshold around
`r example_data$n_for_effect_plot` subjects.

## Parameter Recovery at One Study Size

Power only tells us how often the effect is detected. It does not guarantee the
effect is estimated precisely. The replicate-level estimates below show the
distribution of fitted covariate effects at
`r example_data$n_for_effect_plot` subjects.

```{r fig-effect, fig.width = 6, fig.height = 4}
effect_plot_data <- example_data$effect_plot_data

ggplot(
  effect_plot_data,
  aes(x = factor(study_size), y = estimate)
) +
  geom_boxplot(fill = "#b7d1df", colour = "#2f5c85", width = 0.5) +
  geom_jitter(
    aes(colour = significant),
    width = 0.08,
    height = 0,
    size = 2,
    alpha = 0.8
  ) +
  geom_hline(yintercept = example_data$true_effect, linetype = 2, colour = "#1b7f3b") +
  geom_hline(yintercept = 0, linetype = 3, colour = "#8a3b12") +
  labs(
    x = "Subjects",
    y = "Estimated covariate effect on log(CL)",
    colour = "LRT significant",
    title = "Detection and precision are related but not identical"
  ) +
  theme_minimal()
```

The green dashed line is the true effect used to generate the data, and the
dotted brown line marks zero. Some replicates detect the effect cleanly, while
others remain close to zero or underestimate the magnitude. That is a useful
reminder that a design can be adequately powered for detection before it is
strong for precise estimation.

## Reporting Guidance

When you use SSE for design work, report more than a single power number:

- the assumed true effect size;
- the study-size grid;
- the number of converged replicates;
- the likelihood-ratio threshold used for significance;
- the distribution of fitted covariate effects, not just the pass rate.

This is especially important when convergence problems become non-negligible, or
when Monte Carlo error is large because the number of replicates is small.

## Where To Go Next

Use this vignette as the conceptual starting point, then see
`vignette("runSSE", package = "nlmixr2sse")` for the package's broader
workflow support around restartable runs, alternate parameter sources, and
plotting helpers.
